Multi-wavelength Radio Continuum Emission Studies of Dust-free Red Giants

### 2013-07-13 aj_submission_2/ms.tex
Jul 13: Resubmitted to AJ: If any more edits required then make in new directory

We also include the predicted radio spectrum from the theoretical Alfv\'{e}n wave-driven outflow model for $\alpha$ Tau \citep{1989AcA....39..51K} in Figure \ref{fig:fig2} to demonstrate how radio observations can empirically challenge theoretical models. This model has a fully-ionized outflow inside 10 $R_{\star}$, and has a mass loss rate of  6.3 $\times$ 10$^{-9}$ $M_{\odot}$ yr$^{-1}$, more than two orders of magnitude higher than the more recent estimate given in Table \ref{tab:tab1}. As the radio opacity is proportional to $n _{\rm{e}} n _{\rm{ion}}$, where $n_{\rm{e}}$ and $n_{\rm{ion}}$ are the electron and ion number densities respectively, this model greatly overestimates the actual radio flux density at all VLA wavelengths. The linear Alfv\'{e}n wave models for $\alpha$ Boo \citep{1988AcA....38..107K} also assume full ionization and have higher mass loss rates than the established value given in Table \ref{tab:tab1}, predicting higher flux densities than observed. The lack of agreement between the Alfv\'en wave-driven wind models of \cite{1988AcA....38..107K,1989AcA....39..51K} and our observed radio fluxes may not necessarily be due to an incorrect  wind driving mechanism and instead may be due to the simplifications and  uncertainties in these models, such as wind densities, magnetic field strengths, and damping lengths. For example, the mass loss rate is very sensitive to the radial surface magnetic field strength (i.e., $\dot{M} \propto B^4$) in these Alfv\'en wave models \citep{1983ApJ...275..808H} so a small uncertainty in the mass loss rate can lead to a large uncertainty in the magnetic field strength. Relaxing some of these simplifications such as purely radial flows or non-assumption of the WKB approximation \citep{1995ApJ...454..901C} may also also lead to better agreement with our radio data.

Recently, \cite{2013A&A...553A...3O} has detected a layer of CO in the outer atmosphere of $\alpha$ Tau (i.e. a so-called MOLsphere) which extends out to $\sim$2.5 $R_{\star}$, has a temperature of $1500 \ \pm 200$ K, and a CO column density of $\sim$1$\times 10^{20}$ cm$^{-2}$. They were unable to constrain the geometrical thickness $\Delta L$, of the MOLsphere from the data however, and arbitrarily set it to 0.1 $R_{\star}$. It can be shown that such a MOLsphere would have an optical depth of $\tau _{\rm{6\,cm}} = 4.6$ at C-band and would produce a corresponding flux density of  0.06 mJy which is considerably lower than our high S/N measurement of 0.15 mJy. Here, we have conservatively assumed that the electrons in this region of the atmosphere come from singly ionized metals and have an abundance of $\sim$1$\times 10^{-5} n_{\rm{H}}$, where $n_{\rm{H}}$ is the total hydrogen number density. We have also assumed the CO filling factor to be unity. The disagreement in values between our radio data and the predicted flux for an optically thick disk could mean that the MOLsphere is optically thin at long VLA wavelengths and that the radio emission emanates from the more ionized material closer to the star. It can be shown that the MOLsphere becomes optically thin at C-band (i.e. $\tau _{\rm{6\,cm}} < 1$) for $\Delta L >$0.45 $R_{\star}$, so if the long VLA wavelength radio emission from $\alpha$ Tau comes from a region closer in to the star, then the MOLsphere either has a geometrical width  $>$0.45 $R _{\star}$ or has a filling factor less than unity.

\subsection{Radio Spectral Indices} \label{disc:disc3}
Long wavelength radio emission from non-dusty K spectral-type red giants is due to thermal free-free emission in their partially ionized outflows while shorter wavelength radio emission emanates from nearly static lower atmospheric layers. The radio flux density-frequency relationship for these stars is usually found to be intermediate between that expected from the isothermal stellar disk emission, where $\alpha$ follows the Rayleigh-Jeans tail of the Planck function (i.e., $\alpha = +2$), and that from an optically thin plasma ($\alpha = -0.1$). It can be shown that the expected radio spectrum from a spherically symmetric isothermal outflow with a constant velocity and ionization fraction varies as $\nu ^{0.6}$ \citep{1975MNRAS.170...41W,1975AA....39..217O,1975AA....39....1P}. If we relax some of the assumptions about the outflow in this constant property wind model and instead assume that the electron density and temperature vary as a function of distance from the star $r$, and have the power-law form $n_{e}(r) \propto r^{-p}$ and $T_{e}(r) \propto r^{-n}$ respectively, then

@@

\begin{equation}
\alpha = \frac{4p -6.2 -0.6n}{2p-1-1.35n}
\label{eq:eq1}
\end{equation}
\citep[e.g.,][]{1987ApJ...312..813S}.

@@

Equation 1 can be used in conjunction with our new spectral index for each star to calculate the density and temperature coefficients that may describe their outflows. The combinations of the electron temperature and density coefficients are shown for each star in Figure 5 ($\alpha$ Boo is represented by the solid line and $\alpha$ Tau by the dash-dotted line) along with the coefficients obtained by assuming either an isothermal flow ($n=0$) or a constant velocity flow ($p=2$). One potential explanation for spectral indices of stellar outflows being larger than 0.6 is that the wind is still accelerating in the radio emitting region, if the thermal gradients are assumed to be small. If we ignore thermal gradients completely, then the density coefficients are $p  =2.7$ and 5.5 for $\alpha$ Boo and $\alpha$ Tau, respectively. From mass conservation arguments assuming a steady flow, the power law coefficients for the velocity profiles of each star can be found, ie., $v(r) \propto r^{p-2}$. For $\alpha$ Boo we find $v(r) \propto r^{0.7}$ while for $\alpha$ Tau we find $v(r) \propto r^{3.5}$. This suggests that our long VLA wavelengths may probe a steep acceleration region for $\alpha$ Tau's outflow but for $\alpha$ Boo, may probe a region where the wind is close to its terminal velocity.

@@

The assumption of completely ignoring thermal gradients in a stellar outflow is probably unreliable however. It is likely that some form of Alfv\`{e}n waves are required to lift the material out of the gravitational potential as suggested by \cite{1980ApJ...242..260H}. These waves need damping lengths which are much larger than the density scale height $H_{\star}$ (where $H_{\star} \sim 10 \ H_{\odot}$ for our targets), in order to lift the material out of the gravitational potential but $\lesssim 1 \, R_{\star}$ in order to avoid wind terminal velocities greater than those observed \citep[e.g.][]{1983ApJ...275..808H}. It has also been shown that these waves are expected to produce substantial heating deposition near the base of the wind \citep{1982ApJ...261..279H}. If the long wavelength radio emission from $\alpha$ Tau is indeed emanating from the wind acceleration region, then the dissipation of these Alfv\`{e}n waves may introduce thermal gradients in this region and its velocity profile will not be described by $v(r) \propto r^{3.5}$. Furthermore, diverging flow geometries have been invoked as a more realistic representation of stellar atmospheres \citep{1982ApJ...257..264H,1989A&A...209..327J,2006SSRv..122..181V} and so one could write the continuity equation for a steady flow as 
\begin{equation}
\dot{M} _{\star} = f(r)4\pi r^2 \rho (r) v(r),
\end{equation}
where $f(r)$ is a function describing the divergence from a purely radial flow [i.e., $f(r)=1$], $\dot{M} _{\star}$ is the stellar mass loss rate, and $\rho(r)$ is the gas density at $r$. If the non-radial expansion term can be described by a power-law $f(r) \propto r^s$, where $s>1$ indicates super-radial expansion, then $v(r) \propto r^{3.5-s}$ would describe the velocity profile of $\alpha$ Tau (ignoring thermal gradients and assuming a constant ionization fraction). Further out in the wind, where it has reached its terminal velocity, one would also expect a thermal gradient (but now of opposite sign) due to adiabatic expansion and line cooling. If the long wavelength radio emission emanates from this region of the wind then Equation \ref{eq:eq1} provides us with a direct estimate of the temperature coefficient as we can assume the density coefficient is $p=2$. This may be the case for our long VLA wavelength measurements of $\alpha$ Boo in which case $T_{e}(r) \propto r^{-1.65}$.

@@

The larger mass loss rate of $\alpha$ Boo in comparison to $\alpha$ Tau means that the latter has a substantially smaller effective radius at longer wavelengths, as seen in Figure \ref{fig:fig5}. At 6, 13, and 20 cm the effective radius of $\alpha$ Boo at $\tau _{\rm{rad}}$=1/3 is predicted to be 1.6, 2.8, and 3.7 $R_{\star}$ but is only $\sim$1.2 $R_{\star}$ at 6 and 13 cm for $\alpha$ Tau. \cite{1998ApJ...503..396R} predict that $\alpha$ Tau's wind reaches $\sim$80\% of its terminal velocity by 3 $R_{\star}$, but even our longest-wavelength observations are highly unlikely to sample the wind outside the lower velocity layers closer to the star. For $\alpha$ Boo however, \cite{1985pssl.proc..351D} predicts that the wind has reached its terminal velocity by $\sim$2 $R_{\star}$ so based on this model our longest-wavelength measurements are of the region where the wind has reached a steady terminal velocity. From Figure \ref{fig:fig4}, this implies that the $n_{\rm{e}}$ coefficient is $p=2$ and thus the $T_{\rm{e}}$ coefficient is $n=1.65$. Pure adiabatic expansion cooling with no heat source has $n=1.33$ so additional cooling routes must be operating, possibly due to recombination of H\,\textsc{ii} and/or line cooling. Finally, the wind ionization balance may not have become \textit{frozen-in} in the region of $\alpha$ Boo's wind where the radio emission emanates from. If this is true, then the excess slope of the spectral index could be due to a combination of both cooling and changing ionization fraction. In this scenario the temperature coefficient $n$, would be smaller than our derived value because Equation \ref{eq:eq1} assumes a constant ionization fraction.

@@

To investigate the possibility of the wind undergoing more rapid cooling closer in to the star, we adjusted one of the existing models [referred to as `Model A' in \cite{1985pssl.proc..351D}] to include a temperature power-law falloff of the form
\begin{equation}
T_{e}(r)= T_{e}(r_{1})\left(\frac{r_{1}}{r}\right)^{n},
\label{eq:eq2}
\end{equation}
at some distance $r_{1}$ from the star, and used the temperature coefficient $n=1.65$ obtained from our new VLA data assuming a constant velocity flow (see Figure \ref{fig:fig4}). We introduce the distance $r_{1}$ as the outer limit to ionization processes; at $r > r_{1}$, the ionization fraction is only determined by recombination. To calculate the new electron density in the wind regime where this temperature falloff occurs, we used the analytical expression of \cite{1986ApJ...306..605G} to calculate the hydrogen ionization fraction, $x_{\rm{H\,\textsc{ii}}}=n_{\rm{H\,\textsc{ii}}}/n_{\rm{H}}$, where $n_{\rm{HII}}$ and $n_{\rm{H}}$ are the ionized and total hydrogen number densities, respectively. To do so, we need to make a number of assumptions about the wind properties beyond radius $r_{1}$, namely:

>>>>

We also include the predicted radio spectrum from the theoretical Alfv\'{e}n wave-driven outflow model for $\alpha$ Tau \citep{1989AcA....39..51K} in Figure \ref{fig:fig2} to demonstrate how radio observations can empirically challenge theoretical models. This model has a fully-ionized outflow inside 10 $R_{\star}$, and has a mass loss rate of  6.3 $\times$ 10$^{-9}$ $M_{\odot}$ yr$^{-1}$, more than two orders of magnitude higher than the more recent estimate given in Table \ref{tab:tab1}. As the radio opacity is proportional to $n _{\rm{e}} n _{\rm{ion}}$, where $n_{\rm{e}}$ and $n_{\rm{ion}}$ are the electron and ion number densities respectively, this model greatly overestimates the actual radio flux density at all VLA wavelengths. The linear Alfv\'{e}n wave models for $\alpha$ Boo \citep{1988AcA....38..107K} also assume full ionization and have higher mass loss rates than the value given in Table \ref{tab:tab1}, predicting higher flux densities than observed. The lack of agreement between the Alfv\'en wave-driven wind models of \cite{1988AcA....38..107K,1989AcA....39..51K} and our observed radio fluxes may not necessarily be due to an incorrect  wind driving mechanism and instead may be due to the simplifications and  uncertainties in these models, such as wind densities, magnetic field strengths, damping lengths, and flow geometries close to the star. For example, the mass loss rate is very sensitive to the radial surface magnetic field strength (i.e., $\dot{M} \propto B^4$) in these Alfv\'en wave models \citep{1983ApJ...275..808H} so a small uncertainty in the mass loss rate can lead to a large uncertainty in the magnetic field strength. Relaxing some of these simplifications such as purely radial flows or non-assumption of the WKB approximation \citep{1995ApJ...454..901C} may also also lead to better agreement with our radio data.

Recently, \cite{2013A&A...553A...3O} has detected a layer of CO in the outer atmosphere of $\alpha$ Tau (i.e. a so-called MOLsphere) which extends out to $\sim$2.5 $R_{\star}$, has a temperature of $1500 \ \pm 200$ K, and a CO column density of $\sim$1$\times 10^{20}$ cm$^{-2}$. They were unable to constrain the geometrical thickness $\Delta L$, of the MOLsphere from the data however, and arbitrarily set it to 0.1 $R_{\star}$. It can be shown that such a MOLsphere would have an optical depth of $\tau _{\rm{6\,cm}} = 4.6$ at C-band and would produce a corresponding flux density of  0.06 mJy which is considerably lower than our high S/N measurement of 0.15 mJy. Here, we have conservatively assumed that the electrons in this region of the atmosphere come from singly ionized metals and have an abundance of $\sim$1$\times 10^{-5} n_{\rm{H}}$, where $n_{\rm{H}}$ is the total hydrogen number density. We have also assumed the CO filling factor to be unity. The disagreement in values between our radio data and the predicted flux for an optically thick disk could mean that the MOLsphere is optically thin at long VLA wavelengths and that the radio emission emanates from the more ionized material closer to the star. It can be shown that the MOLsphere becomes optically thin at C-band (i.e. $\tau _{\rm{6\,cm}} < 1$) for $\Delta L >$0.46 $R_{\star}$, so if the long VLA wavelength radio emission from $\alpha$ Tau comes from a region closer in to the star, then the MOLsphere either has a geometrical width  $>$0.46 $R _{\star}$ or has a filling factor less than unity.

\subsection{Radio Spectral Indices} \label{disc:disc3}
Long wavelength radio emission from non-dusty K spectral-type red giants is due to thermal free-free emission in their partially ionized outflows while shorter wavelength radio emission emanates from nearly static lower atmospheric layers. The radio flux density-frequency relationship for these stars is usually found to be intermediate between that expected from the isothermal stellar disk emission, where $\alpha$ follows the Rayleigh-Jeans tail of the Planck function (i.e., $\alpha = +2$), and that from an optically thin plasma ($\alpha = -0.1$). It can be shown that the expected radio spectrum from a spherically symmetric isothermal outflow with a constant velocity and ionization fraction varies as $\nu ^{0.6}$ \citep{1975MNRAS.170...41W,1975AA....39..217O,1975AA....39....1P}. If we relax some of the assumptions about the outflow in this constant property wind model and instead assume that the electron density and temperature vary as a function of distance from the star $r$, and have the power law form $n_{e}(r) \propto r^{-p}$ and $T_{e}(r) \propto r^{-n}$ respectively, then

@@

\begin{equation}
\alpha = \frac{4p -6.2 -0.6n}{2p-1-1.35n}
\label{eq:eq1}
\end{equation}
\citep[e.g.,][]{1987ApJ...312..813S}. These power law approximations are only going to be valid over certain radial ranges of the star's outflow.

@@

Equation 1 can be used in conjunction with our new spectral index for each star to calculate the density and temperature coefficients that may describe their outflows. The combinations of the electron temperature and density coefficients are shown for each star in Figure 5 ($\alpha$ Boo is represented by the solid line and $\alpha$ Tau by the dash-dotted line) along with the coefficients obtained by assuming either an isothermal flow ($n=0$) or a constant velocity flow ($p=2$). One potential explanation for spectral indices of stellar outflows being larger than 0.6 is that the wind is still accelerating in the radio emitting region, if the thermal gradients are assumed to be small. For an isothermal flow, the density coefficients are $p  =2.7$ and 5.5 for $\alpha$ Boo and $\alpha$ Tau, respectively. From mass conservation, and assuming a steady flow, the power law coefficients for the velocity profiles of each star can be found, ie., $v(r) \propto r^{p-2}$. For $\alpha$ Boo we find $v(r) \propto r^{0.7}$ while for $\alpha$ Tau we find $v(r) \propto r^{3.5}$. This suggests that our long VLA wavelengths may probe a steep acceleration region for $\alpha$ Tau's outflow but for $\alpha$ Boo, may probe a region where the wind is close to its terminal velocity.

@@

The assumption of shallow thermal gradients in a stellar outflow is probably unreliable however. It is likely that some form of Alfv\`{e}n waves are required to lift the material out of the gravitational potential as suggested by \cite{1980ApJ...242..260H}. These waves need damping lengths which are much larger than the chromospheric density scale height $H$ (where $H \sim 0.01\ R_{\star}$ for our targets), in order to lift the material out of the gravitational potential but $\lesssim 1 \, R_{\star}$ in order to avoid wind terminal velocities greater than those observed \citep[e.g.][]{1983ApJ...275..808H}. It has also been shown that these waves are expected to produce substantial heating near the base of the wind \citep[e.g.][]{1982ApJ...261..279H}. If the long wavelength radio emission from $\alpha$ Tau is indeed emanating from the wind acceleration region, then the dissipation of these Alfv\`{e}n waves may introduce thermal gradients in this region and its velocity profile will not be described by $v(r) \propto r^{3.5}$. Furthermore, diverging flow geometries have been invoked as a more realistic representation of stellar atmospheres \citep{1982ApJ...257..264H,1989A&A...209..327J,2006SSRv..122..181V} and so one could write the area of a flux tube (normalized to its value at $r_{1}$) as $(r/r_{1})^2f(r)$ where $f(r)$ is a function describing the divergence from a purely radial flow [i.e., when $f(r)=1$].  If the non-radial expansion term can be described by a power law $f(r) \propto r^s$, where $s>1$ indicates super-radial expansion, then $v(r) \propto r^{3.5-s}$ would describe the velocity profile of $\alpha$ Tau (assuming an isothermal flow with a constant ionization fraction). Therefore, including diverging geometries reduces the magnitude of the acceleration. Further out in the wind, where it has reached its terminal velocity, one would also expect a thermal gradient (but now of opposite sign) due to adiabatic expansion and line cooling. If the long wavelength radio emission emanates from this region of the wind then Equation \ref{eq:eq1} provides us with a direct estimate of the temperature coefficient as we can assume the density coefficient is $p=2$. This may be the case for our long VLA wavelength measurements of $\alpha$ Boo in which case $T_{e}(r) \propto r^{-1.65}$.

@@

The larger mass loss rate of $\alpha$ Boo in comparison to $\alpha$ Tau means that the latter has a substantially smaller effective radius at longer wavelengths, as seen in Figure \ref{fig:fig5}. At 6, 13, and 20 cm the effective radius of $\alpha$ Boo at $\tau _{\rm{rad}}$=1/3 is predicted to be 1.6, 2.8, and 3.7 $R_{\star}$ but is only $\sim$1.2 $R_{\star}$ at 6 and 13 cm for $\alpha$ Tau. \cite{1998ApJ...503..396R} predict that $\alpha$ Tau's wind reaches $\sim$80\% of its terminal velocity by 3 $R_{\star}$, but even our longest-wavelength observations are highly unlikely to sample the wind outside the lower velocity layers closer to the star. For $\alpha$ Boo however, \cite{1985pssl.proc..351D} predicts that the wind has reached its terminal velocity by $\sim$2 $R_{\star}$ so based on this model our longest-wavelength measurements are of the region where the wind has reached a steady terminal velocity. From Figure \ref{fig:fig4}, this implies that the $n_{\rm{e}}$ coefficient is $p=2$ and thus the $T_{\rm{e}}$ coefficient is $n=1.65$. Pure adiabatic spherical expansion cooling with no heat source has $n=1.33$ so additional cooling routes must be operating, possibly due to line cooling. Finally, the wind ionization balance may not have become \textit{frozen-in} in the region of $\alpha$ Boo's wind where the radio emission emanates from. If this is true, then the excess slope of the spectral index could be due to a combination of both cooling and changing ionization fraction. In this scenario the temperature coefficient $n$, would be smaller than our derived value because Equation \ref{eq:eq1} assumes a constant ionization fraction.

@@

To investigate the possibility of the wind undergoing more rapid cooling closer in to the star, we adjusted one of the existing models [referred to as `Model A' in \cite{1985pssl.proc..351D}] to include a temperature power law falloff of the form
\begin{equation}
T_{e}(r)= T_{e}(r_{1})\left(\frac{r_{1}}{r}\right)^{n},
\label{eq:eq2}
\end{equation}
at some distance $r_{1}$ from the star, and used the temperature coefficient $n=1.65$ obtained from our new VLA data assuming a constant velocity flow (see Figure \ref{fig:fig4}). We introduce the distance $r_{1}$ as the outer limit to ionization processes; at $r > r_{1}$, the ionization fraction is only determined by recombination. To calculate the new electron density in the wind regime where this temperature falloff occurs, we used the analytical expression of \cite{1986ApJ...306..605G} to calculate the hydrogen ionization fraction, $x_{\rm{H\,\textsc{ii}}}=n_{\rm{H\,\textsc{ii}}}/n_{\rm{H}}$, where $n_{\rm{HII}}$ and $n_{\rm{H}}$ are the ionized and total hydrogen number densities, respectively. To do so, we need to make a number of assumptions about the wind properties beyond radius $r_{1}$, namely:
